\chapter{A Brief History of the Early Universe and Cosmological Phase Transitions}
\label{ch:early-universe}

\section{Thermal History of the Early Universe}

The universe we observe today emerged from an early hot dense state that
has been expanding and cooling ever since. Working backwards in time,
the thermal history is punctuated by a sequence of phase transitions,
each corresponding to the breaking of a symmetry as the temperature fell
below a critical value.

The earliest epoch accessible to our theoretical framework is inflation,
a period of accelerated expansion during which the universe expanded by
a factor of order $10^{26}$ over a time of order $10^{-36}$ to
$10^{-32}$ seconds. At the end of inflation began the reheating phase,
during which the universe returned to a thermal state at the reheating
temperature $T_R$ through the coupling of the inflaton field $\phi$ to
the Standard Model particles.

Subsequent to reheating, the thermal history proceeds through a series
of symmetry breaking transitions. At $T\sim 10^{29}\,\text{K}\sim
	10^{16}\,\text{GeV}$, there may have been a GUT phase transition,
\begin{equation}
	\text{GUT} \;\longrightarrow\; \mathrm{SU}(2)_L\times\mathrm{U}(1),
\end{equation}
though whether this transition occurred and at what scale depends on the
specific GUT model. At $T\sim 10^{15}\,\text{K}\sim 100\,\text{GeV}$,
the electroweak symmetry was broken,
\begin{equation}
	\mathrm{SU}(2)_L\times\mathrm{U}(1)
	\;\longrightarrow\; \mathrm{U}_{\text{em}}(1),
\end{equation}
and at $T\sim 2\cdot 10^{12}\,\text{K}\sim 200\,\text{MeV}$ the QCD
confinement transition took place. Each of these transitions is
described by a scalar field acting as an order parameter that acquires
a vacuum expectation value as the temperature falls through $T_c$.

\section{Phase Transitions and the Effective Potential}

The nature of a cosmological phase transition is determined by the
structure of the finite-temperature effective potential
$V_{\text{eff}}(\bar\varphi, T)$. We illustrate this with the
electroweak phase transition, where the order parameter is the Higgs
field $h$. The tree-level Higgs potential at $T=0$ is
\begin{equation}
	V(h) = -m^2 h^2 + \lambda h^4,
	\qquad m^2 > 0,\quad \lambda > 0,\quad
	m^2\sim(100\,\text{GeV})^2.
\end{equation}

% [FIGURE: Wine-bottle potential at T=0]

At $T\neq 0$, finite-temperature corrections generate a
temperature-dependent mass term,
\begin{equation}
	\Delta V(h,T) = cT^2 h^2 + \cdots,\qquad c > 0,
\end{equation}
so that
\begin{equation}
	V(h,T) = V(h) + \Delta V(h,T).
\end{equation}
The positive $T^2 h^2$ term competes with the negative $-m^2 h^2$ term
and tends to push the minimum back toward $h=0$ at high temperature,
restoring the symmetry.

The character of the transition depends on how the minimum of
$V(h,T)$ evolves with temperature. If the minimum moves smoothly and
continuously from $h\neq 0$ at low $T$ to $h=0$ at high $T$, the
transition is \emph{second order} (or a crossover).

% [FIGURE: Second order phase transition, smooth evolution of minimum]

If instead a potential barrier develops between the two minima at some
temperature, so that the two phases coexist and the field must tunnel
through the barrier to reach the new minimum, the transition is
\emph{first order}.

% [FIGURE: First order phase transition, two degenerate minima at T_c]

The potential barrier in a first-order transition can arise from
higher-order terms in the potential. For example, a cubic term $\sim h^3$
generates a barrier, though for the Standard Model Higgs potential alone
this is not sufficient to produce a first-order transition. There is
also the possibility of \emph{symmetry non-restoration}, which occurs
when $c < 0$, so that the finite-temperature correction deepens the
broken-symmetry minimum rather than restoring it. This situation has
attracted considerable attention in recent years.

% [FIGURE: Symmetry non-restoration, c < 0]

\section{Bubble Nucleation}

In a first-order phase transition, the universe finds itself temporarily
trapped in the false vacuum, the metastable minimum of
$V_{\text{eff}}(\bar\varphi, T)$. The field cannot roll classically to
the true vacuum because of the potential barrier separating the two
phases. Instead, the transition proceeds through the nucleation of
bubbles of the true vacuum within the surrounding false vacuum. Initially,
regions of space where the field has already tunnelled to the true vacuum
appear as isolated bubbles,

% [FIGURE: Bubble of true vacuum in false vacuum background]

which then expand, driven by the pressure difference between the two
phases. New bubbles nucleate, and eventually the expanding bubbles
overlap and collide, filling all of space with the true vacuum and
completing the transition.

This process has two important phenomenological consequences that have
attracted significant attention. The first is electroweak baryogenesis,
in which the out-of-equilibrium dynamics at the bubble wall provides
the conditions necessary to generate the observed baryon asymmetry of
the universe. The second is the production of a stochastic gravitational
wave background from bubble collisions and their interaction with the
primordial plasma. The detailed analysis of both of these consequences
requires a quantitative understanding of the bubble nucleation rate and
bubble wall dynamics, which we develop in Part~\ref{part:false-vacuum}.

