\chapter{Toy Model: One-Loop Effective Potential and Renormalization}
\label{ch:toy-model}

We illustrate the general formalism of the previous chapter with an
explicit computation in the simplest setting: a single real scalar field
with Lagrangian
\begin{equation}
	\mathcal{L}(x) = \frac{1}{2}\,\partial_\mu\phi(x)\,\partial^\mu\phi(x)
	- V(\phi(x)),
	\label{eq:real_scalar_field}
\end{equation}
and classical potential
\begin{equation}
	V(\phi) = \frac{\alpha_0}{4!}\,\phi^4 - \frac{m_0^2}{2}\,\phi^2.
\end{equation}
We work in units $\hbar = c = 1$ throughout.

\section{The Classical Field and the Effective Action}

The classical minimum of $V(\phi)$ sits at some $\phi_0$, but the vacuum
expectation value $\langle\phi\rangle$ of the quantum field need not
coincide with it. It is defined by
\begin{equation}
	\langle\phi\rangle \equiv \lim_{J\to 0}
	\frac{\langle 0^+|\phi_{\text{op}}|0^-\rangle_J}
	{\langle 0^+|0^-\rangle_J},
\end{equation}
where the generating functional $W[J]$ encodes the full quantum theory via
\begin{equation}
	\exp\frac{i}{\hbar}W[J]
	= \mathcal{N}\int(\mathcal{D}\phi)
	\exp\frac{i}{\hbar}\left\{S[\phi] + (J,\phi)\right\},
\end{equation}
with $S[\phi] = \int d^4x\,\mathcal{L}(x)$ and
$(J,\phi) = \int d^4x\,J(x)\,\phi(x)$.
The classical field, the VEV in the presence of an external source $J(x)$,
is
\begin{equation}
	\phi_c(x) \equiv
	\frac{\langle 0^+|\phi_{\text{op}}(x)|0^-\rangle_J}
	{\langle 0^+|0^-\rangle_J}
	= \frac{\delta W[J]}{\delta J(x)},
	\label{eq:phi_c}
\end{equation}
and $\langle\phi\rangle$ is recovered as $\lim_{J\to 0}\phi_c(x)$. The
loop expansion is an expansion in powers of $\hbar\,c\,\alpha$, but it
performs an infinite resummation of Feynman graphs rather than a
perturbative truncation.

The natural question is: given a desired profile $\phi_c$, what source
$J$ produces it? To answer this, we trade $J$ for $\phi_c$ as the
independent variable via the Legendre transform,
\begin{equation}
	\Gamma[\phi_c] \equiv W[J] - (J,\phi_c),
	\label{eq:Gamma_def}
\end{equation}
where $J$ is eliminated in favour of $\phi_c$ by inverting
\eqref{eq:phi_c}. The functional $\Gamma[\phi_c]$ is the effective action,
so named because it is a functional of the classical field $\phi_c$ and
thus plays the same role for the quantum theory as $S[\phi]$ does
classically. Differentiating \eqref{eq:Gamma_def} and using
\eqref{eq:phi_c},
\begin{align}
	\frac{\delta\Gamma[\phi_c]}{\delta\phi_c(x)}
	 & = \frac{\delta W[J]}{\delta\phi_c(x)} - J(x)
	- \int d^4y\,\frac{\delta J(y)}{\delta\phi_c(x)}\,\phi_c(y), \notag \\
	\frac{\delta W[J]}{\delta\phi_c(x)}
	 & = \int d^4y\,\frac{\delta W[J]}{\delta J(y)}
	\frac{\delta J(y)}{\delta\phi_c(x)}
	= \int d^4y\,\frac{\delta J(y)}{\delta\phi_c(x)}\,\phi_c(y),
\end{align}
so that
\begin{equation}
	\frac{\delta\Gamma[\phi_c]}{\delta\phi_c(x)} = -J(x).
	\label{eq:eff_action_eom}
\end{equation}
At $J=0$, translational invariance forces $\phi_c$ to be a constant, so
$\langle\phi\rangle$ is the root of
\begin{equation}
	\frac{d\Gamma[\phi_c]}{d\phi_c}\bigg|_{\phi_c=\langle\phi\rangle} = 0.
\end{equation}

\section{The Effective Action as a Generator of 1PI Green's Functions}

Expanding $\Gamma[\phi_c]$ in powers of $\phi_c$,
\begin{equation}
	\Gamma[\phi_c] = \sum_{n=0}^{\infty} \frac{1}{n!}
	\int d^4x_1\cdots d^4x_n\,
	\Gamma_n(x_1,\ldots,x_n)\,\phi_c(x_1)\cdots\phi_c(x_n),
	\label{eq:gamma_expansion}
\end{equation}
we claim that $\Gamma_n$ is the $n$-point 1PI Green function, the sum
of all connected 1PI Feynman graphs with $n$ external legs. To see this,
consider the functional integral
\begin{equation}
	\mathcal{N}\int(\mathcal{D}\phi)
	\exp\frac{i}{\hbar}\left\{\Gamma[\phi] + (J,\phi)\right\}.
\end{equation}
In the limit $\hbar\to 0$ the integrand is dominated by its saddle point,
which occurs at $\phi = \phi_c$ since \eqref{eq:eff_action_eom} is
precisely the saddle-point condition. Thus
\begin{equation}
	\mathcal{N}\int(\mathcal{D}\phi)
	\exp\frac{i}{\hbar}\left\{\Gamma[\phi] + (J,\phi)\right\}
	\xrightarrow{\hbar\to 0}
	\mathcal{N}\exp\frac{i}{\hbar}
	\left\{\Gamma[\phi_c] + (J,\phi_c)\right\},
\end{equation}
and by \eqref{eq:Gamma_def} the right-hand side equals
$\exp\frac{i}{\hbar}W[J]$, so
\begin{equation}
	\exp\frac{i}{\hbar}W[J]
	\xrightarrow{\hbar\to 0}
	\mathcal{N}\int(\mathcal{D}\phi)
	\exp\frac{i}{\hbar}\left\{\Gamma[\phi] + (J,\phi)\right\}.
\end{equation}
In this limit the right-hand side is the sum of all tree graphs of a
theory whose action is $\Gamma[\phi]$, with vertices given by the
$\Gamma_n$ of \eqref{eq:gamma_expansion}. Since $W[J]$ generates all
connected Green functions of the original theory, and any connected
graph decomposes into 1PI components, it follows that $\Gamma_n$ is a
1PI Green function of the original theory. \qed

\section{The Effective Potential}

Alternatively, $\Gamma[\phi_c]$ may be expanded in derivatives of
$\phi_c$,
\begin{equation}
	\Gamma[\phi_c] = \int d^4x\left[
		-U(\phi_c(x))
		+ \frac{1}{2}\,\partial_\mu\phi_c(x)\,\partial^\mu\phi_c(x)\,
		Z(\phi_c(x))
		+ \cdots\right],
\end{equation}
where $U(\phi_c)$ is the effective potential. For a constant configuration
$\phi_c(x) = \rho$ this reduces to
\begin{equation}
	\Gamma[\rho] = -\Omega\,U(\rho),
\end{equation}
where $\Omega$ is the total spacetime volume. To connect $U(\rho)$ to the
1PI Green functions, we pass to momentum space. The Fourier transforms
of $\phi_c$ and $\Gamma_n$ are defined by
\begin{align}
	\tilde{\phi}_c(k) & = \int d^4x\,e^{ik\cdot x}\,\phi_c(x),                          \\
	\phi_c(x)         & = \int\frac{d^4k}{(2\pi)^4}\,e^{-ik\cdot x}\,\tilde{\phi}_c(k),
\end{align}
\begin{equation}
	\tilde{\Gamma}_n(k_1,\ldots,k_n)\,(2\pi)^4\,\delta^4(k_1+\cdots+k_n)
	= \int d^4x_1\cdots d^4x_n\,\Gamma_n(x_1,\ldots,x_n)\,
	e^{i(k_1 x_1+\cdots+k_n x_n)}.
\end{equation}
Substituting these into \eqref{eq:gamma_expansion},
\begin{multline}
	\Gamma[\phi_c]
	= \sum_{n=0}^{\infty}\frac{1}{n!}
	\int d^4x_1\cdots d^4x_n\,
	\Gamma_n(x_1,\ldots,x_n)\,\phi_c(x_1)\cdots\phi_c(x_n) \\
	= \sum_{n=0}^{\infty}\frac{1}{n!}
	\int\frac{d^4k_1}{(2\pi)^4}\cdots\frac{d^4k_n}{(2\pi)^4}
	\int d^4x_1\cdots d^4x_n\,
	\Gamma_n(x_1,\ldots,x_n)\,
	e^{-i(k_1 x_1+\cdots+k_n x_n)}
	\,\tilde{\phi}(k_1)\cdots\tilde{\phi}(k_n) \\
	= \sum_{n=0}^{\infty}\frac{1}{n!}
	\int\frac{d^4k_1}{(2\pi)^4}\cdots\frac{d^4k_n}{(2\pi)^4}\,
	(2\pi)^4\,\delta^4(k_1+\cdots+k_n)\,
	\tilde{\Gamma}_n(k_1,\ldots,k_n)\,
	\tilde{\phi}(k_1)\cdots\tilde{\phi}(k_n).
	\label{eq:rewriting_Gamma}
\end{multline}
Setting $\phi_c(x) = \rho$ gives $\tilde{\phi}_c(k) =
	(2\pi)^4\delta^4(k)\,\rho$, so each momentum integral collapses,
\begin{multline}
	\Gamma[\rho]
	= \sum_{n=0}^{\infty}\frac{1}{n!}
	\int\frac{d^4k_1}{(2\pi)^4}\cdots\frac{d^4k_n}{(2\pi)^4}\,
	(2\pi)^4\,\delta^4(k_1+\cdots+k_n)\,
	\tilde{\Gamma}_n(k_1,\ldots,k_n) \\
	\cdot\,(2\pi)^4\delta^4(k_1)\,\rho\cdots(2\pi)^4\delta^4(k_n)\,\rho \\
	= \sum_{n=0}^{\infty}\frac{1}{n!}
	\underbrace{(2\pi)^4\delta^4(0)}_{\Omega}\,
	\rho^n\,\tilde{\Gamma}_n(0,\ldots,0),
\end{multline}
and therefore
\begin{equation}
	\Gamma[\rho] = \Omega\sum_{n=0}^{\infty}
	\frac{\rho^n}{n!}\,\tilde{\Gamma}_n(0),
	\label{eq:Gamma_rho}
\end{equation}
where $\tilde{\Gamma}_n(0)\equiv\tilde{\Gamma}_n(0,\ldots,0)$ and
$\Omega = (2\pi)^4\delta^4(0)$ is the total spacetime volume. Comparing
\eqref{eq:Gamma_rho} with $\Gamma[\rho]=-\Omega\,U(\rho)$,
\begin{equation}
	U(\rho) = -\sum_{n=0}^{\infty}\frac{\rho^n}{n!}\,\tilde{\Gamma}_n(0),
\end{equation}
which is simply the statement that the Taylor coefficients of $U$ are the
zero-momentum 1PI Green functions,
\begin{equation}
	-\frac{d^n U(\rho)}{d\rho^n}\bigg|_{\rho=0} = \tilde{\Gamma}_n(0).
\end{equation}
The effective potential is thus entirely determined by the 1PI Green
functions of the original theory evaluated at vanishing external momenta.
In the next section we compute it explicitly at one loop.

\section{Spontaneous Symmetry Breaking and the Shifted Field}

When spontaneous symmetry breaking occurs, it is more natural to work
with the shifted field
\begin{equation}
	\eta(x) = \phi(x) - \langle\phi\rangle,
\end{equation}
so that the effective potential in terms of the shifted variable becomes
\begin{equation}
	U\!\left(\rho - \langle\phi\rangle\right) = -\sum_{n=0}^{\infty}
	\frac{\left[\rho - \langle\phi\rangle\right]^n}{n!}\,
	\widetilde{\Gamma}_n^{(s)}(0),
\end{equation}
where $\widetilde{\Gamma}_n^{(s)}(0)$ are the 1PI Green functions defined
with respect to $\eta(x)$. The physically relevant quantities extracted
from $U(\rho)$ at the minimum $\rho = \langle\phi\rangle$ are
\begin{equation}
	\frac{dU(\rho)}{d\rho}\bigg|_{\rho=\langle\phi\rangle} = 0,
\end{equation}
which determines $\langle\phi\rangle$, together with the renormalized mass
and coupling constant,
\begin{equation}
	\frac{d^2 U(\rho)}{d\rho^2}\bigg|_{\rho=\langle\phi\rangle} = m^2,
	\qquad
	\frac{d^4 U(\rho)}{d\rho^4}\bigg|_{\rho=\langle\phi\rangle} = \alpha.
\end{equation}
Since $m^2 \geq 0$, these conditions mean that $\langle\phi\rangle$ sits
at a minimum of $U(\rho)$, in direct analogy with $\rho_0$ being a minimum
of the classical potential $V(\rho)$. The parameters $m^2$ and $\alpha$
are the free parameters of the renormalized theory.

\section{One-Loop Effective Potential}

We now compute $U(\rho)$ to one-loop order via saddle-point integration.
Starting from
\begin{equation}
	\exp\frac{i}{\hbar}W[J] = \mathcal{N}\int\mathcal{D}\phi\;
	\exp\frac{i}{\hbar}\left\{S[\phi] + (J,\phi)\right\},
\end{equation}
the saddle point occurs at $\phi = \phi_0$, defined by
\begin{equation}
	\frac{\delta S[\phi]}{\delta\phi(x)}\bigg|_{\phi=\phi_0} = -J(x).
\end{equation}
Here $\phi_0(x)$ is a functional of $J$, and as $J\to 0$ it satisfies the
classical equation of motion. Expanding $S[\phi]$ about $\phi_0$,
\begin{equation}
	\begin{split}
		S[\phi_0 + \phi_c] = S[\phi_0]
		 & + \int d^4x\;\phi_c(x)
		\left[\frac{\delta S[\phi]}{\delta\phi}\right]_{\phi=\phi_0} \\
		 & + \frac{1}{2}\int d^4x\,d^4y\;\phi_c(x)\,\phi_c(y)
		\left[\frac{\delta^2 S[\phi]}{\delta\phi(x)\,\delta\phi(y)}
			\right]_{\phi=\phi_0}   \\
		 & + S_2[\phi_c,\phi_0],
	\end{split}
\end{equation}
where $S_2$ collects all higher-order terms. Introducing the propagator
function
\begin{equation}
	\langle x|\,i\Delta^{-1}[\phi_0]\,|y\rangle
	\equiv \frac{\delta^2 S[\phi]}{\delta\phi(x)\,\delta\phi(y)}
	\bigg|_{\phi=\phi_0},
	\label{eq:propagator_function_def}
\end{equation}
the expansion becomes
\begin{equation}
	S[\phi_0+\phi_c] = S[\phi_0] - (J,\phi_c)
	+ \frac{1}{2}\left(\phi_c,\,i\Delta^{-1}[\phi_0]\,\phi_c\right)
	+ S_2[\phi_c,\phi_0].
\end{equation}
Substituting into the path integral and shifting $\phi\to\phi+\phi_0$,
\begin{multline}
	\exp\frac{i}{\hbar}W[J]
	= \mathcal{N}\int(\mathcal{D}\phi)
	\exp\frac{i}{\hbar}\left\{S[\phi+\phi_0]+(\phi+\phi_0,J)\right\} \\
	= \mathcal{N}\int(\mathcal{D}\phi)
	\exp\frac{i}{\hbar}\left\{S[\phi_0]
	+ \frac{1}{2}(\phi,\,i\Delta^{-1}[\phi_0]\,\phi)
	+ S_2[\phi,\phi_0] + (\phi_0,J)\right\},
\end{multline}
which factors as
\begin{multline}
	= \exp\frac{i}{\hbar}\left\{S[\phi_0]+(\phi_0,J)\right\}
	\cdot\left\{\mathcal{N}\int(\mathcal{D}\phi)
	\exp\frac{i}{2\hbar}(\phi,\,i\Delta^{-1}[\phi_0]\,\phi)\right\}
	\cdot\left\langle\exp\frac{i}{\hbar}S_2\right\rangle,
\end{multline}
where the Gaussian integral evaluates to
$\left\{\det i\Delta^{-1}[\phi_0]\right\}^{-1/2}$ and $\langle\,\cdot\,\rangle$
denotes the functional average with respect to the Gaussian weight
$\exp\frac{i}{2\hbar}(\phi,\,i\Delta^{-1}[\phi_0]\,\phi)$. Taking the
logarithm,
\begin{equation}
	W[J] = S[\phi_0] + (\phi_0,J)
	+ \frac{i\hbar}{2}\ln\det\left\{i\Delta^{-1}[\phi_0]\right\}
	+ W_2[J],
	\label{eq:W_expansion}
\end{equation}
where $W_2[J] = -i\hbar\ln\langle\exp\frac{i}{\hbar}S_2\rangle$.

To determine the order of $W_2[J]$, rescale $\phi = \hbar^{1/2}\tilde\phi$,
\begin{align}
	\hbar^{-1}S_2[\phi,\phi_0]
	 & = \hbar^{-1}\phi\phi\phi
	\left[\frac{\delta^3 S}{\delta\phi^3}\right]_{\phi=\phi_0}
	+ \cdots \notag                                 \\
	 & = \hbar^{-1/2}\tilde\phi\tilde\phi\tilde\phi
	\left[\frac{\delta^3 S}{\delta\phi^3}\right]_{\phi=\phi_0}
	+ \cdots,
\end{align}
so that
\begin{align}
	W_2[J] & = -i\hbar\ln\left[1+\mathcal{O}(\hbar^{1/2})\right]
	\sim \mathcal{O}(\hbar^{3/2}).
\end{align}
Since the loop expansion of $W[J]$ contains only integer powers of $\hbar$,
the fractional contribution must actually vanish and $W_2[J]\sim\mathcal{O}(\hbar^2)$.
Thus the first two terms in \eqref{eq:W_expansion} are the tree-level
approximation and the third term is the complete one-loop correction,
\begin{equation}
	W[J] = S[\phi_0] + (\phi_0,J)
	+ \frac{i\hbar}{2}\ln\det\left\{i\Delta^{-1}[\phi_0]\right\}
	+ \mathcal{O}(\hbar^2).
\end{equation}

To obtain $\Gamma[\phi_c] = W[J]-(J,\phi_c)$, we express $S[\phi_0]$
as a functional of $\phi_c$. At tree level $\phi_c = \phi_0$, so
$\phi_c - \phi_0 \equiv \phi_1 \sim \mathcal{O}(\hbar)$,
\begin{equation}
	\phi_c(x) \equiv \frac{\delta W[J]}{\delta J(x)}
	= \phi_0(x) + \phi_1(x).
\end{equation}
Expanding $S[\phi_0] = S[\phi_c-\phi_1]$ about $\phi_c$,
\begin{align}
	S[\phi_0] & = S[\phi_c]
	- \int d^4x\,\phi_1(x)
	\left\{\frac{\delta S[\phi_0]}{\delta\phi}\right\}_{\phi=\phi_c}
	+ \mathcal{O}(\hbar^2) \notag                                \\
	          & = S[\phi_c] + (\phi_1,J) + \mathcal{O}(\hbar^2).
\end{align}
Substituting into $\Gamma[\phi_c]$,
\begin{align}
	\Gamma[\phi_c]
	 & = S[\phi_c] + \underbrace{(\phi_1,J)}_{\mathcal{O}(\hbar)}
	+ (\phi_c,J)
	+ \frac{i\hbar}{2}\ln\det\left\{i\Delta^{-1}[\phi_c]\right\}
	+ \mathcal{O}(\hbar^2) \notag                                 \\
	 & = S[\phi_c] + (\phi_c,J)
	+ \frac{i\hbar}{2}\ln\det\left\{i\Delta^{-1}[\phi_c]\right\}
	+ \mathcal{O}(\hbar^2),
\end{align}
where the $(\phi_1,J)$ term is absorbed into $\mathcal{O}(\hbar^2)$.
Setting $J=0$ and $\phi_c(x)=\rho$ (const), and using
$\Gamma[\rho]=-\Omega\,U(\rho)$ and $S[\rho]=-\Omega\,V(\rho)$,
\begin{equation}
	-\Omega\,U(\rho) = -\Omega\,V(\rho)
	+ \frac{i\hbar}{2}\ln\det\left\{i\Delta^{-1}[\rho]\right\}
	+ \mathcal{O}(\hbar^2),
\end{equation}
\begin{equation}
	\boxed{U(\rho) = V(\rho)
		- \frac{i\hbar}{2\Omega}\ln\det\left\{i\Delta^{-1}[\rho]\right\}
		+ \mathcal{O}(\hbar^2).}
	\label{eq:U_rho}
\end{equation}

The determinant is evaluated using $\ln\det = \mathrm{Tr}\ln$,
\begin{align}
	\ln\det\left\{i\Delta^{-1}[\rho]\right\}
	 & = \mathrm{Tr}\ln\left\{i\Delta^{-1}[\rho]\right\} \notag \\
	 & = \int\frac{d^4k}{(2\pi)^4}
	\ln\langle k|\,i\Delta^{-1}[\rho]\,|k\rangle,
\end{align}
so that
\begin{equation}
	U(\rho) = V(\rho)
	- \frac{i\hbar}{2}\int\frac{d^4k}{(2\pi)^4}\,
	\Omega^{-1}\ln\langle k|\,i\Delta^{-1}[\rho]\,|k\rangle
	+ \mathcal{O}(\hbar^2).
	\label{eq:U_10.41}
\end{equation}
To evaluate $\Omega^{-1}\langle k|\,i\Delta^{-1}[\rho]\,|k\rangle$, expand
$\phi(x) = \rho + \phi_1(x)$ with $\rho$ constant and
$\left.\partial V/\partial\phi\right|_{\phi=\rho}=0$,
\begin{align}
	S[\phi] & = \int d^4x\left[\frac{1}{2}\,\partial_\mu\phi_1\,
		                     \partial^\mu\phi_1 - V(\phi_1+\rho)\right] \notag \\
	        & = -\frac{1}{2}\int d^4x\;\phi_1
	\left[\Box + V''(\rho)\right]\phi_1 + \cdots.
\end{align}
By \eqref{eq:propagator_function_def},
\begin{align}
	\langle x|\,i\Delta^{-1}[\rho]\,|y\rangle
	 & = \frac{\delta^2 S[\phi]}{\delta\phi_1(x)\,\delta\phi_1(y)}
	\bigg|_{\phi_1=0} \notag                                       \\
	 & = -\left[\Box + V''(\rho)\right]\delta^4(x-y).
\end{align}
Taking the matrix element between momentum eigenstates (see
Appendix~\ref{app:matrix-element}),
\begin{equation}
	\Omega^{-1}\langle k|\,i\Delta^{-1}[\rho]\,|k\rangle
	= k^2 - V''(\rho).
\end{equation}
Substituting into \eqref{eq:U_10.41} and rotating to Euclidean momentum
space via $k^0\to ik^0_E$, $k^2\to -k_E^2$, $d^4k = i\,d^4k_E$,
\begin{equation}
	U(\rho) = V(\rho)
	+ \frac{\hbar}{2}\int\frac{d^4k_E}{(2\pi)^4}
	\ln\left[k_E^2 + V''(\rho)\right]
	+ \mathcal{O}(\hbar^2),
	\label{eq:U_euclidean}
\end{equation}
where the $i\pi$ contribution from $\ln(-k_E^2-V'')$ is a field-independent
constant and has been dropped. The momentum integral is evaluated in
Appendix~\ref{app:oneloop-integral}, with the result
\begin{multline}
	\frac{\hbar}{2}\int\frac{d^4k_E}{(2\pi)^4}\ln(k_E^2+V'')
	= \hbar\left[\frac{\Lambda^2}{32\pi^2}V''(\rho)
		+ \frac{[V''(\rho)]^2}{64\pi^2}
		\left(-\frac{1}{2}+\ln\frac{V''(\rho)}{\Lambda^2}\right)
		+ \mathcal{O}\!\left(\frac{1}{\Lambda^2}\right)\right].
\end{multline}
We can therefore write
\begin{equation}
	U(\rho) = V(\rho) + \hbar\,V_1(\rho) + \mathcal{O}(\hbar^2),
\end{equation}
where the one-loop correction is
\begin{equation}
	V_1(\rho) = \frac{\Lambda^2}{32\pi^2}V''(\rho)
	+ \frac{[V''(\rho)]^2}{64\pi^2}
	\left[-\frac{1}{2}+\ln\frac{V''(\rho)}{\Lambda^2}\right].
	\label{eq:V1}
\end{equation}
To separate the divergent from the convergent part, introduce an arbitrary
renormalization scale $\mu$ via
\begin{equation}
	\ln\frac{V''}{\Lambda^2}
	= \ln\frac{V''}{\mu^2} + \ln\frac{\mu^2}{\Lambda^2},
\end{equation}
so that
\begin{multline}
	V_1(\rho)
	= \frac{\Lambda^2}{32\pi^2}V''(\rho)
	+ \frac{(V''(\rho))^2}{64\pi^2}
	\left[-\frac{1}{2}+\ln\frac{V''}{\mu^2}+\ln\frac{\mu^2}{\Lambda^2}
		\right] \\
	= \underbrace{\frac{1}{32\pi^2}\left[\Lambda^2 V''(\rho)
			- \frac{1}{2}[V''(\rho)]^2\ln\frac{\Lambda^2}{\mu^2}
			\right]}_{\text{divergent}}
	+ \underbrace{\frac{(V'')^2}{64\pi^2}
		\left[-\frac{1}{2}+\ln\frac{V''}{\mu^2}\right]}_{\text{convergent}}.
\end{multline}
The divergent part must be removed by renormalization, which we turn to
next.


\section{Renormalization}

The theory is renormalizable if the counterterms needed to cancel the
divergences are of the same form as terms already present in the
Lagrangian. For a classical potential $V(\rho)$ that is a polynomial of
degree $n$, the second derivative $V''(\rho)$ is of degree $n-2$, and the
one-loop correction $V_1(\rho)$ in \eqref{eq:V1} is of degree $2(n-2) =
2n-4$ from the $(V'')^2$ term. Renormalizability then requires
\begin{equation}
    2n - 4 \leq n \quad\Rightarrow\quad n \leq 4,
\end{equation}
a condition satisfied by our choice of $V(\rho)$.

The bare parameters are split into renormalized parameters and
counterterms,
\begin{align}
    \alpha_0 &= \alpha_1 + \delta\alpha, \\
    m_0^2 &= m_1^2 + \delta m^2,
\end{align}
where $\delta\alpha$ and $\delta m^2$ are divergent in perturbation theory
but $\alpha_1$ and $m_1^2$ are required to be finite. Since the
counterterms only arise at the quantum level, $\delta\alpha,\,\delta m^2
\sim \mathcal{O}(\hbar)$. The Lagrangian becomes
\begin{equation}
    \mathcal{L}(x) = \frac{1}{2}\,\partial_\mu\phi\,\partial^\mu\phi
    - V(\phi)
    + \underbrace{\frac{\delta m^2}{2}\phi^2
    - \frac{\delta\alpha}{4!}\phi^4}_{\text{counterterms}},
\end{equation}
with the renormalized classical potential
\begin{equation}
    V(\phi) = \frac{\alpha_1}{4!}\phi^4 - \frac{m_1^2}{4}\phi^2.
\end{equation}
The classical vacuum value $\rho_0$ is found from
\begin{align}
    \frac{dV(\phi)}{d\phi} &= \frac{\alpha_1}{3!}\phi^3
    - \frac{m_1^2}{2}\phi = 0 \notag\\
    &\Rightarrow \phi\left(\frac{\alpha_1}{3}\phi^2 - m_1^2\right) = 0,
\end{align}
giving $\phi = 0$ or
\begin{equation}
    \rho_0 = m_1\left(\frac{3}{\alpha_1}\right)^{\frac{1}{2}}.
\end{equation}
Since the counterterms are of order $\hbar$, they contribute at the same
order as the one-loop correction and the full effective potential is
\begin{equation}
    U(\rho) = V(\rho) + \hbar\,V_1(\rho)
    + \frac{\delta\alpha}{4!}\rho^4 - \frac{\delta m^2}{2}\rho^2.
\end{equation}
Choosing $\delta\alpha$ and $\delta m^2$ to precisely cancel the divergent
part of $V_1(\rho)$,
\begin{align}
    \delta\alpha &= \frac{3\alpha_1^2\hbar}{32\pi^2}
    \ln\frac{\Lambda^2}{\mu^2}, \\
    \delta m^2 &= \frac{\alpha_1\hbar}{32\pi^2}
    \left(\Lambda^2 + \frac{m_1^2}{2}\ln\frac{\Lambda^2}{\mu^2}\right),
\end{align}
where the right-hand sides are ambiguous up to finite additive terms, the
effective potential becomes
\begin{equation}
    U(\rho) = V(\rho) + \frac{\hbar}{64\pi^2}\left[V''(\rho)\right]^2
    \left[-\frac{1}{2} + \ln\frac{V''(\rho)}{\mu^2}\right].
    \label{eq:effective_potential}
\end{equation}

\section{Physical Parameters from the Effective Potential}

We now extract the VEV, the renormalized mass, and the renormalized
coupling constant from \eqref{eq:effective_potential}. Setting $k \equiv
\hbar/(32\pi^2)$ for brevity, the successive derivatives of $U(\rho)$ are
computed using $V = \frac{\alpha_1}{4!}\rho^4 - \frac{m_1^2}{4}\rho^2$,
so $V' = \frac{\alpha_1}{3!}\rho^3 - \frac{m_1^2}{2}\rho$,
$V'' = \frac{\alpha_1}{2}\rho^2 - \frac{m_1^2}{2}$,
$V''' = \alpha_1\rho$, $V'''' = \alpha_1$, $V''''' = 0$.

Differentiating \eqref{eq:effective_potential},
\begin{multline}
    U' = V' + \frac{\hbar}{64\pi^2}\left[2V''V'''\!\left(-\frac{1}{2}
    + \ln\frac{V''}{\mu^2}\right)
    + (V'')^2\cdot\frac{1}{V''}\cdot\frac{V'''}{\mu^2}\cdot\mu^2\right] \\
    = V' + \frac{\hbar}{64\pi^2}\left[-V''V''' + 2V''V'''\ln\frac{V''}{\mu^2}
    + V''V'''\right],
\end{multline}
\begin{equation}
    \Rightarrow U' = V' + k\,V''V'''\ln\frac{V''}{\mu^2}.
    \label{eq:Uprime}
\end{equation}
Differentiating again,
\begin{multline}
    U'' = V'' + k\left[(V''')^2\ln\frac{V''}{\mu^2}
    + V''V''''\ln\frac{V''}{\mu^2}
    + V''V'''\cdot\frac{1}{V''}\cdot\frac{V'''}{\mu^2}\cdot\mu^2\right],
\end{multline}
\begin{equation}
    \Rightarrow U'' = V'' + k\left[\left\{(V''')^2
    + \alpha_1 V''\right\}\ln\frac{V''}{\mu^2} + (V''')^2\right].
\end{equation}
The third derivative,
\begin{multline}
    U''' = V''' + k\Bigg[\left\{(V''')^2 + \alpha_1 V''\right\}'
    \ln\frac{V''}{\mu^2} \\
    + \left\{(V''')^2 + \alpha_1 V''\right\}
    \left(\ln\frac{V''}{\mu^2}\right)' + 2V'''V''''\Bigg] \\
    = V''' + k\left[\left(2V'''V'''' + \alpha_1 V'''\right)
    \ln\frac{V''}{\mu^2}
    + \left\{(V''')^2 + \alpha_1 V''\right\}\frac{V'''}{V''}
    + 2\alpha_1 V'''\right],
\end{multline}
\begin{equation}
    \Rightarrow U''' = V''' + k\left[3\alpha_1 V'''\ln\frac{V''}{\mu^2}
    + \frac{(V''')^3}{V''} + 3\alpha_1 V'''\right].
    \label{eq:Uprimeprimeprime}
\end{equation}
The fourth derivative,
\begin{multline}
    U'''' = V'''' + k\left[3\alpha_1 V''''\ln\frac{V''}{\mu^2}
    + 3\alpha_1 V'''\cdot\frac{1}{V''}\cdot\frac{V'''}{\mu^2}\cdot\mu^2
    \right. \\
    \left. + \frac{3(V''')^2 V''''}{V''}
    - \frac{(V''')^3}{(V'')^2}\cdot V''' + 3\alpha_1 V''''\right] \\
    = \alpha_1 + k\left[3\alpha_1^2\ln\frac{V''}{\mu^2}
    + 3\alpha_1\frac{(V''')^2}{V''}
    + \frac{3\alpha_1(V''')^2}{V''}
    - \frac{(V''')^4}{(V'')^2} + 3\alpha_1^2\right] \\
    = \alpha_1 + k\left[3\alpha_1^2\ln\frac{V''}{\mu^2}
    + \frac{6\alpha_1(V''')^2}{V''}
    - \frac{(V''')^4}{(V'')^2} + 3\alpha_1^2\right],
\end{multline}
\begin{equation}
    \Rightarrow U'''' = \alpha_1 + k\alpha_1^2\left[3\ln\frac{V''}{\mu^2}
    + \frac{6\alpha_1\rho^2}{V''}
    - \frac{\alpha_1^2\rho^4}{(V'')^2} + 3\right].
    \label{eq:Uprrrr}
\end{equation}

\subsection*{The VEV}

The minimum $\langle\phi\rangle = \rho_0 + \rho_1$ is found by expanding
$U'(\rho_0+\rho_1)=0$ to first order,
\begin{equation}
    U'(\rho_0) + \rho_1\,U''(\rho_0) = 0
    \quad\Rightarrow\quad
    \rho_1 = -\frac{U'(\rho_0)}{U''(\rho_0)}.
\end{equation}
We first evaluate $V''(\rho_0)$ and $V'''(\rho_0)$ at the classical
minimum $\rho_0^2 = 3m_1^2/\alpha_1$,
\begin{align}
    V''(\rho_0) &= \frac{\alpha_1}{2}\rho_0^2 - \frac{m_1^2}{2}
    = \frac{\alpha_1}{2}\cdot\frac{3m_1^2}{\alpha_1} - \frac{m_1^2}{2}
    = \frac{3m_1^2}{2} - \frac{m_1^2}{2} = m_1^2,
\end{align}
\begin{equation}
    \Rightarrow V''(\rho_0) = m_1^2.
\end{equation}
Also, $V'(\rho_0) = 0$ by definition of $\rho_0$, so from
\eqref{eq:Uprime},
\begin{equation}
    U'(\rho_0) = k\,m_1^2\,\alpha_1\,\rho_0\ln\frac{m_1^2}{\mu^2}.
\end{equation}
For $U''(\rho_0)$, the $\mathcal{O}(\alpha_1^2)$ terms are dropped,
\begin{align}
    U''(\rho_0) &= m_1^2 + k\left[(\alpha_1^2\rho_0^2
    + \alpha_1 m_1^2)\ln\frac{m_1^2}{\mu^2}
    + \alpha_1^2\rho_0^2\right] \notag\\
    &= m_1^2 + k\,\alpha_1 m_1^2\ln\frac{m_1^2}{\mu^2}
    + \mathcal{O}(\alpha_1^2),
\end{align}
\begin{equation}
    \Rightarrow U''(\rho_0) = m_1^2\left(1
    + \alpha_1 k\ln\frac{m_1^2}{\mu^2}\right)
    + \mathcal{O}(\alpha_1^2).
\end{equation}
Hence,
\begin{align}
    \rho_1 &= -\frac{k\,m_1^2\,\alpha_1\,\rho_0\ln\frac{m_1^2}{\mu^2}}
    {m_1^2\left(1 + \alpha_1 k\ln\frac{m_1^2}{\mu^2}\right)} \notag\\
    &= -k\,\alpha_1\,\rho_0\ln\frac{m_1^2}{\mu^2}
    \left(1 + \alpha_1 k\ln\frac{m_1^2}{\mu^2}\right)^{-1} \notag\\
    &= -k\alpha_1\rho_0\ln\frac{m_1^2}{\mu^2}
    \left[1 - \alpha_1 k\ln\frac{m_1^2}{\mu^2}
    + \mathcal{O}(\alpha_1^2)\right] \notag\\
    &= -k\alpha_1\rho_0\ln\frac{m_1^2}{\mu^2} + \mathcal{O}(\alpha_1^2),
\end{align}
\begin{equation}
    \frac{\rho_1}{\rho_0} = -\frac{\hbar\alpha_1}{32\pi^2}
    \ln\frac{m_1^2}{\mu^2} + \mathcal{O}(\alpha_1^2).
\end{equation}
Setting $\hbar=1$, the full VEV is
\begin{equation}
    \boxed{\langle\phi\rangle = m_1\left(\frac{3}{\alpha_1}\right)^{\frac{1}{2}}
    \left[1 - \frac{\alpha_1}{32\pi^2}\ln\frac{m_1^2}{\mu^2}
    + \mathcal{O}(\alpha_1^2)\right].}
\end{equation}

\subsection*{The Renormalized Mass}

The renormalized mass is $m^2 \equiv U''(\rho_0+\rho_1)$. Expanding,
\begin{equation}
    m^2 = U''(\rho_0) + \rho_1\,U'''(\rho_0). \tag{I}
\end{equation}
From \eqref{eq:Uprimeprimeprime}, dropping $\mathcal{O}(\alpha_1^2)$
terms,
\begin{align}
    U'''(\rho_0) &= \alpha_1\rho_0 + k\left[3\alpha_1^2\rho_0
    \ln\frac{m_1^2}{\mu^2}
    + \frac{(\alpha_1\rho_0)^3}{m_1^2}
    + 3\alpha_1^2\rho_0\right] \notag\\
    &= \alpha_1\rho_0 + \mathcal{O}(\alpha_1^2), \tag{II}
\end{align}
\begin{equation}
    U'''(\rho_0) = \alpha_1\rho_0
    + k\left[3\alpha_1^2\rho_0\ln\frac{m_1^2}{\mu^2}\right]. \tag{III}
\end{equation}
Also,
\begin{equation}
    U''(\rho_0) = m_1^2 + k\left[(\alpha_1^2\rho_0^2
    + \alpha_1 m_1^2)\ln\frac{m_1^2}{\mu^2}
    + \alpha_1^2\rho_0^2\right].
\end{equation}
Substituting (II) and (III) into (I), using
$\rho_1 = -k\alpha_1\rho_0\ln\frac{m_1^2}{\mu^2}$ and
$\rho_1 U'''(\rho_0) = -k\alpha_1\rho_0\ln\frac{m_1^2}{\mu^2}
\cdot 3m_1^2 + \mathcal{O}(\alpha_1^2)$,
\begin{align}
    m^2 &= m_1^2\left(1 + 4\alpha_1 k\ln\frac{m_1^2}{\mu^2}
    + 3\alpha_1 k\right)
    - 3m_1^2\alpha_1 k\ln\frac{m_1^2}{\mu^2}
    + \mathcal{O}(\alpha_1^2) \notag\\
    &= m_1^2\left[1 + 4\alpha_1 k\ln\frac{m_1^2}{\mu^2} + 3\alpha_1 k
    - 3\alpha_1 k\ln\frac{m_1^2}{\mu^2}\right]
    + \mathcal{O}(\alpha_1^2) \notag\\
    &= m_1^2\left[1 + \alpha_1 k\ln\frac{m_1^2}{\mu^2}
    + 3\alpha_1 k\right] + \mathcal{O}(\alpha_1^2) \notag\\
    &= m_1^2\left[1 + 3\alpha_1 k\left(1
    + \frac{1}{3}\ln\frac{m_1^2}{\mu^2}\right)\right]
    + \mathcal{O}(\alpha_1^2),
\end{align}
\begin{equation}
    \Rightarrow\quad\boxed{m^2 = m_1^2\left[1
    + \frac{3\alpha_1}{32\pi^2}\left(1
    + \frac{1}{3}\ln\frac{m_1^2}{\mu^2}\right)
    + \mathcal{O}(\alpha_1^2)\right].}
\end{equation}

\subsection*{The Renormalized Coupling Constant}

The renormalized coupling is $\alpha \equiv U''''(\rho_0+\rho_1)$.
Evaluating \eqref{eq:Uprrrr} at $\rho=\rho_0$ and keeping terms to
$\mathcal{O}(\alpha_1)$,
\begin{align}
    \alpha &= \alpha_1 + k\alpha_1^2\left[3\ln\frac{V''}{\mu^2}
    + \frac{6\alpha_1\rho^2}{V''}
    - \frac{\alpha_1^2\rho^4}{(V'')^2} + 3\right] \notag\\
    &= \alpha_1\left[1 + 3\alpha_1 k\left(1
    + \ln\frac{V''}{\mu^2}\right)
    + \mathcal{O}(\alpha_1^2)\right],
\end{align}
\begin{equation}
    \Rightarrow\quad\boxed{\alpha = \alpha_1\left[1
    + \frac{3\alpha_1}{32\pi^2}\left(1
    + \ln\frac{m_1^2}{\mu^2}\right)
    + \mathcal{O}(\alpha_1^2)\right].}
\end{equation}

\section{The Massless Case and the Running Coupling Constant}

We now consider the massless limit $m_1=0$, in which
\begin{align}
    V' &= \frac{\alpha_1}{6}\rho^3, &
    V'' &= \frac{\alpha_1\rho^2}{2}.
\end{align}
From \eqref{eq:Uprrrr},
\begin{align}
    U''''(\rho) &= \alpha_1 + k\alpha_1^2\left[
    3\ln\frac{\alpha_1\rho^2}{2\mu^2}
    + \frac{6\alpha_1\rho^2}{\frac{\alpha_1\rho^2}{2}}
    - \frac{\alpha_1^2\rho^4}{\frac{\alpha_1^2\rho^4}{4}} + 3\right]
    \notag\\
    &= \alpha_1 + k\alpha_1^2\left[
    3\ln\frac{\alpha_1\rho^2}{2\mu^2} + 12 - 4 + 3\right],
\end{align}
\begin{equation}
    \Rightarrow U''''(\rho) = \alpha_1 + k\alpha_1^2\left[11
    + 3\ln\frac{\alpha_1\rho^2}{2\mu^2}\right].
\end{equation}
Setting $\alpha_1 \equiv \alpha$ and defining the running coupling
constant $\alpha(\rho) \equiv U''''(\rho)$,
\begin{equation}
    \alpha(\rho) = \alpha + \frac{k\alpha^2}{1}\left[11
    + 3\ln\frac{\alpha\rho^2}{2\mu^2}\right].
\end{equation}
We introduce a new scale $\mu_0$ defined by the renormalization condition
$U''''(\mu_0) = \alpha$, which gives
\begin{align}
    \alpha &= \alpha + k\alpha^2\left[11
    + 3\ln\frac{\alpha\mu_0^2}{2\mu^2}\right] \notag\\
    \Rightarrow\quad\ln\frac{\alpha\mu_0^2}{2\mu^2} &= -\frac{11}{3}.
\end{align}
Using this to eliminate the explicit $\ln(\alpha/2\mu^2)$ dependence,
\begin{align}
    \alpha(\rho) &= \alpha + k\alpha^2\left[3\cdot\frac{11}{3}
    + 3\ln\frac{\alpha\rho^2}{2\mu^2}\right] \notag\\
    &= \alpha + k\alpha^2\left[-3\ln\frac{\alpha\mu_0^2}{2\mu^2}
    + 3\ln\frac{\alpha\rho^2}{2\mu^2}\right] \notag\\
    &= \alpha + 3k\alpha^2\ln\!\left(
    \frac{\frac{\alpha\rho^2}{2\mu^2}}
    {\frac{\alpha\mu_0^2}{2\mu^2}}\right),
\end{align}
\begin{equation}
    \Rightarrow\quad\boxed{\alpha(\rho) = \alpha
    + \frac{3k\alpha^2}{32\pi^2}\ln\frac{\rho^2}{\mu_0^2},}
\end{equation}
where $\alpha = \alpha(\mu_0)$ is the value of the running coupling at
$\rho=\mu_0$. Inverting,
\begin{align}
    \frac{1}{\alpha(\rho)}
    &= \frac{1}{\alpha(\mu_0) + \frac{3k\alpha^2(\mu_0)}{32\pi^2}
    \ln\frac{\rho^2}{\mu_0^2}}
    = \frac{1}{\alpha(\mu_0)}\left[1
    + \frac{3k\alpha(\mu_0)}{32\pi^2}
    \ln\frac{\rho^2}{\mu_0^2}\right]^{-1},
\end{align}
\begin{equation}
    \Rightarrow\quad\frac{1}{\alpha(\rho)}
    = \frac{1}{\alpha(\mu_0)}\left(1
    - \frac{3k\alpha(\mu_0)}{32\pi^2}
    \ln\frac{\rho^2}{\mu_0^2}\right)
    + \text{higher order.}
\end{equation}

\section{Spontaneous Symmetry Breaking by Radiative Corrections?}

A natural question arises: can SSB occur in the massless theory purely
due to radiative corrections, generating a nonzero $\langle\phi\rangle$
dynamically? From \eqref{eq:Uprime} with $m_1=0$,
\begin{align}
    U'(\rho) &= V' + k\,V''V'''\ln\frac{V''}{\mu^2} \notag\\
    &= \frac{\alpha}{6}\rho^3 + k\,\frac{\alpha\rho^2}{2}\cdot\alpha\rho
    \cdot\ln\frac{\alpha\rho^2}{2\mu^2} \notag\\
    &= \frac{\rho^3}{6}\left[\alpha
    + 3k\alpha^2\ln\frac{\alpha\rho^2}{2\mu^2}\right].
\end{align}
Writing $\ln\frac{\alpha\rho^2}{2\mu^2}
= \ln\frac{\rho^2}{\mu^2} + \ln\frac{\alpha}{2}$, where the last term is
a field-independent constant,
\begin{equation}
    U'(\rho) = \frac{\rho^3}{6}\left[\alpha
    + \frac{3\hbar\alpha^2}{32\pi^2}
    \left(\ln\frac{\rho^2}{\mu^2}
    + \text{const}\right)\right].
    \label{eq:first_derivative_potential}
\end{equation}
Setting $U'(\langle\phi\rangle)=0$ for a nontrivial root gives
\begin{align}
    &\alpha + \frac{3k\alpha^2}{32\pi^2}
    \ln\frac{\langle\phi\rangle^2}{\mu^2} = 0 \notag\\
    \Rightarrow\quad
    &\alpha\left(1 + \frac{3k\alpha}{32\pi^2}
    \ln\frac{\langle\phi\rangle^2}{\mu^2}\right) = 0 \notag\\
    \Rightarrow\quad
    &\hbar\alpha\ln\frac{\langle\phi\rangle^2}{\mu^2}
    = -\frac{32\pi^2}{3} + \mathcal{O}(\hbar).
    \label{eq:misleading_root}
\end{align}
This root must be rejected. The loop expansion is valid only when
$\hbar\alpha$ is small, but \eqref{eq:misleading_root} requires
\begin{equation}
    \hbar\alpha\ln\frac{\langle\phi\rangle^2}{\mu^2}
    \sim \mathcal{O}(1),
\end{equation}
meaning higher-order terms are not small and cannot be neglected. The
minimum lies outside the domain of validity of the one-loop approximation. Let us explain this in slightly more details. The loop expansion is an expansion in powers of $\hbar$, with the
$L$-loop contribution carrying a factor of $\hbar^L$. The one-loop
approximation is therefore valid when the one-loop correction is small
compared to the tree-level term,
\begin{equation}
    \frac{\hbar V_1(\rho)}{V(\rho)} \ll 1,
\end{equation}
or more precisely in this case, when
\begin{equation}
    \hbar\alpha\ln\frac{\langle\phi\rangle^2}{\mu^2} \ll 1.
\end{equation}
The spurious root requires precisely the opposite,
$\hbar\alpha\ln(\langle\phi\rangle^2/\mu^2) \sim \mathcal{O}(1)$, so that
the logarithm is of order $1/(\hbar\alpha)$ and $\langle\phi\rangle$ sits
exponentially far from $\mu$. At such field values the two-loop
contribution goes as
\begin{equation}
    (\hbar\alpha)^2\cdot\left(\ln\frac{\langle\phi\rangle^2}{\mu^2}
    \right)^2 \sim \mathcal{O}(1),
\end{equation}
and similarly for all higher loops, so every order contributes equally and
truncating at one loop is unjustified. The root is therefore rejected, not

of validity of the approximation.



The technical reason for this failure is that the two terms in
\eqref{eq:first_derivative_potential} are of order $\alpha$ and
$\alpha^2$ respectively, and cannot cancel each other within the range of
validity of perturbation theory: an $\mathcal{O}(\alpha)$ quantity cannot
cancel an $\mathcal{O}(\alpha^2)$ quantity. However, if the model were
modified so that $\alpha$ and $\alpha^2$ were replaced by two independent
coupling constants, the cancellation could occur and one would obtain a
genuine theory of SSB by radiative corrections. The simplest such model
is massless scalar electrodynamics, to which we turn in the next section.
